Cooling Requirements for the Vertical Shear Instability, \\ with Implications for Protoplanetary Disks

### 2015-04-08 paper.tex
minor notation update
a punchier title and reworded abstract (with ultimately very similar content).  I think these versions are less technical and more inviting to the average arXiv reader, but still loyal to the work.   We can discuss iterations.

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, vertical shear is weak in thin disks, 
  so the VSI can easily be stabilized by buoyancy forces in the
  presence of a stable vertical entropy  gradient. % , which is also
  % expected in astrophysical disks. 
  Consequently,  the VSI
  requires a rather short thermal or cooling timescale in
  sub-adabatically stratified disks.      
  A quantitative examination of this thermodynamic 
  requirement is needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. % We explore how
  % the VSI depends on the thermodynamic response of the disk by
  % varying $\beta$.
  We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. Thus, in a thin disk the thermal timescale must be shorter than 
  the dyamical timescale by $O(\epsilon)\ll 1$ for the VSI to operate.  
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI.

@@

Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height, as evident in
Fig. \ref{eqm_structure}. Eq. \ref{bcrit_est} then suggest that with
increasing height, buoyancy would eventually stabilize the VSI (since 
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$). 
% This means that 
% disturbances localized away from the disk midplane are easily stabilized by
% buoyancy. 
This argument assumes vertically localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

%it would imply you can always have VSI sufficiently close to the
%midplane

>>>>

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\def \OmK {\Omega_{\rm K}}

@@

\title{Cooling Requirements for the Vertical Shear Instability, \\
 with Implications for Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that operate in warmer disks are less effective. 
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport and accretion flows in cold protoplanetary disks.  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong stabilizing influence in cold disks, where 
  heating is dominated by external irradiation.  Rapid cooling, via radiative losses, reduces the effective buoyancy
  and allows the VSI to operate.  In this paper, we quantify the cooling timescale, $t_c$, needed for growth of the VSI
  and consider the implications for protoplanetary disks where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in vertically global and radially local disk models.
  For irradiated disks, we find that the VSI is most vigorous for rapid cooling with $t_c < \Omega_K^{-1} \epsilon |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect ratio, $\epsilon \ll 1$; the radial 
  powerlaw temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer cooling times, the VSI 
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula (MMSN) disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  Growth is suppressed by long cooling times both in the opaque 
  inner disk and the optically thin outer disk.  A reduction in the dust opacity by a factor of 10 increases cooling times 
  enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.

%

%  A robust hydrodynamic mechanism for turbulent angular momentum
%  transport is important for cold accretion disks such as 
%  protoplanetary disks. The vertical shear instability (VSI) has 
%  recently been suggested to provide such a pathway, since it operates in
%  disks with rotation profiles that depend on the height from the 
%  disk midplane, which is generally expected in  
%  astrophysical disks. However, vertical shear is weak in thin disks, 
%  so the VSI can easily be stabilized by buoyancy forces in the
%  presence of a stable vertical entropy  gradient. % , which is also
%  % expected in astrophysical disks. 
%  Consequently,  the VSI
%  requires a rather short thermal or cooling timescale in
%  sub-adabatically stratified disks.      
%  A quantitative examination of this thermodynamic 
%  requirement is needed to assess the viability of
%  the VSI in realistic disks. To this end,  
%  we study the linear, axisymmetric VSI in radially local, 
%  vertically global models of stably stratified accretion disks in which the disk  
%  temperature $T$ is relaxed to its initial value over a characteristic timescale 
%  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
%  and $\Omega_k$ is the Keplerian rotation frequency. % We explore how
%  % the VSI depends on the thermodynamic response of the disk by
%  % varying $\beta$.
%  We find that 
%  increasing $\beta$ even slightly away from    
%  zero dramatically reduce VSI growth rates. 
%  For vertically isothermal disks, we quantify the
%  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
%  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
%  $\gamma$ is the adiabatic index. Thus, in a thin disk the thermal timescale must be shorter than 
%  the dyamical timescale by $O(\epsilon)\ll 1$ for the VSI to operate.  
%  We apply our results  
%  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
%  estimate for the thermal timescale $\beta$ in the MMSN based on dust
%  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
%  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
%  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
%  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
%  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
%  characteristic growth time of $\sim 30$ local orbital periods.  
%  % ,and the dominant VSI mode in the MMSN is the `fundamental
%  % corrugation mode' which causes the disk column to oscillate
%  % vertically.
%  We thus suggest the VSI is at least dynamically important in the 
%  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI. 
Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height. Eq. \ref{bcrit_est} then suggest that with
increasing height, buoyancy would eventually stabilize the VSI (since
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$).
% This means that 
% disturbances localized away from the disk midplane are easily stabilized by
% buoyancy. 
This argument assumes vertically localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

@@
